### 2016-04-19 src/3_TimeSeries.tex
avance capitulo 2. Non-stationary definition falta

\section{Characteristics of Financial Time Series}
%\section{Stylized facts of asset returns}
\label{sec:stylizedfacts}

@@

Returns of a time series $y_t$ is defined as:
\begin{equation}
\label{eq:returndef}
r_t = y_t - y_{t-1}  \, ,
\end{equation}

@@

\begin{description}
\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant. The ACF measures the linear predictability of the series at time $t$ using only the value at time $s$:

\begin{equation}
\rho(s,t) = \frac{\gamma(s,t)}{\sqrt{\gamma(s,s)\gamma(t,t)}}
\end{equation}
\noindent where $\gamma(s,t)$ is the autocovariance function measures the linear dependence between two points on the same series observed at different times, it is defined as the second moment product:
\begin{equation}
\gamma_y(s,t) = E[(y_s-\mu_s)(y_t - \mu_t)]
\end{equation}
However, the autocorrelation in the absolute and squared returns is always positive,
significant and decays slowly. In addition, the autocorrelation in the absolute
returns is generally higher than the autocorrelation in the corresponding
squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
 \ref{fig:returnacf}. The figure shows how the correlations are significant even
 for very long lags, this imply that a long-memory process because of its
 long-range dependence. 
 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. Figure
\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
in average Monday has negative returns. However, this is changing in time,
figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
three months and it shows that Mondays and Wednesdays has positive returns in
average.

\begin{figure}[h!]
\centering
\begin{subfigure}[b]{0.45\textwidth}
 \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
 \caption{Average daily return: 1928-2007}
 \label{fig:mondayeffect}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
 \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
 \caption{Average daily return: 2007}
 \label{fig:mondayeffect2}
\end{subfigure}
\end{figure}
\end{description}

\subsection{Stationary }
A strictly stationary times series $Y_t$ is one for which the probabilistic behaviour
of every collection of values $\{Y_{t_1},Y_{t_2},\dots,Y_{t_L}\}$ is identical
to that of the time shifted set, more precisely: \[ P\{Y_{t_1} \leq
c_1,\dots,Y_{t_L} \leq c_L\} = P\{Y_{t_1+h} \leq c_1,\dots,Y_{t_L+h} \leq c_L\}
\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
assess it from a single data set. The weak version of this definition imposes
conditions only on the two first two moments.

A weakly stationary time series is a process which mean, variance and auto
covariance do not change over time: \begin{eqnarray*} E(Y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(Y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}

\noindent with $\lambda(s,t) = E[(Y_s-\mu)(Y_t - \mu)]$

>>>>

\section{Characteristics of Financial Time Series}
There are several characteristics and concepts of financial time series and here we will discuss the more important ones.
\subsection{Stylized facts of asset returns}
\label{sec:stylizedfacts}

@@

@@

\begin{description}
\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant. Returns of a time series $y_t$ is defined as $r_t = y_t - y_{t-1}$. The ACF measures the linear predictability of the time series $y_t$ at time $t$ using only the value at time $s$. ACF is defined as:
\begin{equation*}
\rho(s,t) = \frac{\gamma(s,t)}{\sqrt{\gamma(s,s)\gamma(t,t)}}
\end{equation*}
\noindent where $\gamma(s,t)$ is the auto-covariance function that measures the linear dependence between two points on the same series observed at different times. It is defined as the second moment product:
\begin{equation*}
\gamma_y(s,t) = E[(y_s-\mu_s)(y_t - \mu_t)]
\end{equation*}

The ACF in the absolute and squared returns is always positive,
significant and decays slowly. In addition, the ACF in the absolute
returns is generally higher than the ACF in the corresponding
squared returns.

Figure \ref{fig:returnacf} shows the ACF of the returns of SPY stock where the correlations are significant even
 for very long lags, this implies a long-memory process.

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. 
\end{description}

\subsection{Stationary }
A strictly stationary times series $y_t$ is one for which the probabilistic behaviour
of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_L\}
\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
assess it from a single data set. The weak version of this definition imposes
conditions only on the two first two moments.

A weakly stationary time series is a process which mean, variance and auto
covariance do not change over time: \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}

\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$
